Introducing ICBe: Very High Recall and Precision Event Extraction from Narratives about International Crises

### 2022-09-30 replication_paper/political_analysis_draft/ICBe_political_analysis_submission_rmd.tex
pa submission

\begin{abstract}
How do international crises unfold? We conceptualize international
relations as a strategic chess game between adversaries and develop a
systematic way to measure pieces, moves, and gambits accurately and
consistently over a hundred years of history. We introduce a new
ontology and dataset of international events called ICBe based on a very
high-quality corpus of narratives from the International Crisis Behavior
(ICB) Project. We demonstrate that ICBe has higher coverage, recall, and
precision than existing state of the art datasets and conduct two
detailed case studies of the Cuban Missile Crisis (1962) and the
Crimea-Donbas Crisis (2014). We further introduce two new event
visualizations (event iconography and crisis maps), an automated
benchmark for measuring event recall using natural language processing
(synthetic narratives), and an ontology reconstruction task for
objectively measuring event precision. We make the data, supplementary
appendix, replication material, and visualizations of every historical
episode available at a companion website www.crisisevents.org.
\end{abstract}

@@

If we could record every important interaction between countries in all
of diplomacy, military conflict, and international political economy,
how much unique information would this chronicle amount to, and how
surprised would we be to see something new? In other words, what is the
entropy of international relations? This record could in principle be
unbounded, but the central conceit of social science is that there are
structural regularities that limit what actors can do, their best
options, and even which actors are likely to survive (Brecher 1999;
Reiter 2015). If so, then these events can be systematically measured,
and accordingly, massive effort is expended in social science attempting
to record these regularities.\footnote{See work on crises (Brecher and
  Wilkenfeld 1982; Beardsley et al. 2020), militarized disputes (Palmer
  et al. 2021; Gibler 2018), wars (Sarkees and Wayman 2010; Reiter,
  Stam, and Horowitz 2016), organized violence (Ralph Sundberg and Mihai
  Croicu 2016; Pettersson and Eck 2018), political violence (Raleigh et
  al. 2010), sanctions (Felbermayr et al. 2020), and international
  agreements (Kinne 2020; Owsiak, Cuttner, and Buck 2018), dispute
  resolution (Frederick, Hensel, and Macaulay 2017), and diplomacy
  (Moyer, Turner, and Meisel 2020; Sechser 2011).} Thanks to
improvements in natural language processing, more open-ended efforts
have begun to capture entire unstructured streams of events from
international news reports.\footnote{See Li et al. (2021); Beieler et
  al. (2016); Brandt et al. (2018); Boschee et al. (2015). On event
  extraction from images and social media see Zhang and Pan (2019) and
  Steinert-Threlkeld (2019).} How close these efforts are to accurately
measuring all or even most of what is essential in international
relations is an open empirical question, one for which we provide new
evidence here.

Our contribution is a high coverage ontology and event dataset for key
historical episodes in 20th and 21st-century international relations. We
develop a large, flexible ontology of international events with the help
of both human coders and natural language processing. We apply it
sentence-by-sentence to an unusually high-quality corpus of historical
narratives of international crises (Brecher 1999; Brecher, James, and
Wilkenfeld 2000; Wilkenfeld and Brecher 2000; Brecher et al. 2021). The
result is a new lower bound estimate of how much actually happens
between states during pivotal historical episodes. We then develop
several methods for objectively gauging how well these event codings
reconstruct the information contained in the original narrative. We
conclude by benchmarking our event codings against several current state
of the art event data collection efforts. We find that existing systems
produce sequences of events that do not contain enough information to
reconstruct the underlying historical episode. The underlying
fine-grained variation in international affairs is unrecognizable
through the lens of current quantification efforts.

@@

The second axis (columns) compares the very high ontological coverage of
ICBe to existing state of the art systems in production and with global
coverage. They begin with our contribution ICBe, alongside other
event-level datasets including CAMEO dictionary lookup-based systems
(Historical Phoenix (Althaus et al. 2019); ICEWS (Boschee et al. 2015;
Hegre et al. 2020); Terrier (Grant et al. 2017)), the Militarized
Interstate Disputes Incidents dataset, and the UCDP-GED dataset (Ralph
Sundberg and Mihai Croicu 2016; Pettersson and Eck 2018; Sundberg and
Melander 2013).\footnote{Other related datasets that insufficiently
  overlap ICBe's domain for comparison include BCOW (Leng and Singer
  1988), WEIS (McClelland 1978), CREON (Hermann 1984), CASCON
  (Bloomfield and Moulton 1989), SHERFACS (Sherman 2000), Real-Time
  Phoenix (Brandt et al. 2018), and COfEE (Balali, Asadpour, and Jafari
  2021) (see histories in Merritt (1994) and Schrodt and Hall (2006)).}
The final set of columns compares episode-level datasets beginning with
the original ICB project (Brecher et al. 2021; Brecher and Wilkenfeld
1982; Beardsley et al. 2020); the Militarized Interstate Disputes
dataset (Palmer et al. 2021; Gibler 2018), and the Correlates of War
(Sarkees and Wayman 2010). With the exception of large-scale CAMEO
dictionary-based systems, the existing state of the art quantitative
datasets ignore the vast majority of the information content found in
international relations.\footnote{See Balali, Asadpour, and Jafari
  (2021) for a recent review of ontological depth and availability of
  Gold Standard example text.}

@@

We find both the visualizations of macrostructure and head-to-head
comparisons of ICBe codings to the raw text to strongly support the
quality of ICBe, but as with recall, we seek a more objective detached
universal benchmark. Our proposed measure is a reconstruction task to
see whether our intended ontology can be recovered through only
unsupervised clustering of sentences they were applied to. Figure 4
shows the location of every sentence from the ICBe corpus in semantic
space as embedded using the same large language model as before, and the
median location of each ICBe event tag applied to those
sentences.\footnote{We preprocess sentences to replace named entities
  with a generic Entity token.} Labels reflect the individual leaves of
the ontology and colors reflect the higher level coarse branch nodes of
the ontology. If ICBe has high precision, substantively similar tags
ought to have been applied to substantively similar source text, which
is what we see both in two dimensions in the main plot and via
hierarchical clustering on all dimensions in the dendrogram along the
righthand side.\footnote{Hierchcial clustering on cosine similarity and
  with Ward's method.}

@@

We take the same approach here in Figure 5, selecting four specific
CAMEO event codings and checking how often they reflect a true
real-world event from the Crimea-Donbas synthetic narrative. We choose
four event types around key moments in the crisis. The start of the
crisis revolves around Ukraine backing out of a trade deal with the EU
in favor of Russia, but ``sign formal agreement'' events act more like a
topic detector with dozens of events generated by discussions of a
possible agreement but not the actual agreement which never
materialized. The switch is caught by the ``reject plan, agreement to
settle dispute'', but also continues for Victor Yanukovych even after he
was removed from power because of articles retroactively discussing the
cause of his removal. Events for ``use conventional military force''
capture a threshold around the start of hostilities and who the
participants were but not any particular battles or campaigns. Likewise,
``impose embargo, boycott, or sanctions'' captures the start of waves of
sanctions and from who but are effectively constant as the news coverage
does not distinguish between subtle changes or additions. In sum,
dictionary-based methods on news corpora tend to have high recall
because they parse everything in the news, but for the same reason,
their specificity for most event types is too low to back out individual
chess-like sequencing that ICBe aims to record.

@@

We release several open-source products along with supporting code and
documentation to further advance the study of IR, event extraction, and
natural language processing. The first is the International Crisis
Behavior Events (ICBe) dataset, an event-level aggregation of what took
place during the crises identified by the ICB project. These data are
appropriate for statistical analysis of hard questions about the
sequencing of events (e.g.~escalation and de-escalation of
conflicts).\footnote{Using ICBe data Gannon (2022) finds that
  cross-domain crises are shorter and less violent than same-domain
  crises.} Second, we provide a coder-level disaggregation with multiple
codings of each sentence by experts and undergrads that allows for the
introduction of uncertainty and human interpretation of events. Further,
we release a direct mapping from the codings to the source text at the
sentence level as a new resource for natural language processing.
Finally, we provide a companion website that incorporates detailed
visualizations of all of the data introduced here at
www.crisisevents.org.

@@

We thank the ICB Project and its directors and contributors for their
foundational work and their help with this effort. We make special
acknowledgment of Michael Brecher for helping found the ICB project in
1975, creating a resource that continues to spark new insights to this
day. We thank the many undergraduate coders for their patience and
dedication. Thanks to the Center for Peace and Security Studies and its
membership for comments. Special thanks to Rebecca Cordell, Philip
Schrodt, Zachary Steinert-Threlkeld, and Zhanna Terechshenko for their
generous feedback. Thank you to the cPASS research assistants that
contributed to this project: Helen Chung, Daman Heer, Syeda ShahBano
Ijaz, Anthony Limon, Erin Ling, Ari Michelson, Prithviraj Pahwa, Gianna
Pedro, Tobias Stodiek, Yiyi `Effie' Sun, Erin Werner, Lisa Yen, and
Ruixuan Zhang.

@@

This article's data, supplementary appendix, replication material, and
visualizations of every historical episode are available on the GitHub
repository
\href{https://urldefense.com/v3/__https://github.com/CenterForPeaceAndSecurityStudies/ICBEventData__;!!Mih3wA!WxDJtEczKfxGTh0S2Krunap8ReymFEL5iTWaSfOHeqlSdyfRx77zmjBSWO1OAm13$}{ICBEventData}
and through the companion website www.crisisevents.org.

@@

\leavevmode\vadjust pre{\hypertarget{ref-brecherEscalationWarTwentieth2000}{}}%
Brecher, Michael, Patrick James, and Jonathan Wilkenfeld. 2000.
{``Escalation and War in the Twentieth Century: {Findings} from the
{International Crisis Behavior Project}.''} \emph{What Do We Know about
War}, 37--53.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-hegreIntroducingUCDPCandidate2020}{}}%
Hegre, Håvard, Mihai Croicu, Kristine Eck, and Stina Högbladh. 2020.
{``Introducing the {UCDP Candidate Events Dataset}.''} \emph{Research \&
Politics} 7 (3): 2053168020935257.

@@

\leavevmode\vadjust pre{\hypertarget{ref-petterssonOrganizedViolence19892018}{}}%
Pettersson, Therése, and Kristine Eck. 2018. {``Organized Violence,
1989\textendash 2017.''} \emph{Journal of Peace Research} 55 (4):
535--47. \url{https://doi.org/10.1177/0022343318784101}.

>>>>

\begin{abstract}
How do international crises unfold? We conceptualize international
relations as a strategic chess game between adversaries and develop a
systematic way to measure pieces, moves, and gambits accurately and
consistently over a hundred years of history. We introduce a new
ontology and dataset of international events called ICBe based on a very
high-quality corpus of narratives from the International Crisis Behavior
(ICB) Project. We demonstrate that ICBe has higher coverage, recall, and
precision than existing state of the art datasets and conduct two
detailed case studies of the Cuban Missile Crisis (1962) and the
Crimea-Donbas Crisis (2014). We further introduce two new event
visualizations (event iconography and crisis maps), an automated
benchmark for measuring event recall using natural language processing
(synthetic narratives), and an ontology reconstruction task for
objectively measuring event precision. We make the data, supplementary
appendix, replication material, and visualizations of every historical
episode available at a companion website crisisevents.org.
\end{abstract}

@@

If we could record every important interaction between countries in all
of diplomacy, military conflict, and international political economy,
how much unique information would this chronicle amount to, and how
surprised would we be to see something new? In other words, what is the
entropy of international relations? This record could in principle be
unbounded, but the central conceit of social science is that there are
structural regularities that limit what actors can do, their best
options, and even which actors are likely to survive (Brecher 1999;
Reiter 2015). If so, then these events can be systematically measured,
and accordingly, massive effort is expended in social science attempting
to record these regularities.\footnote{See work on crises (Brecher and
  Wilkenfeld 1982; Beardsley et al. 2020), militarized disputes (Palmer
  et al. 2021; Gibler 2018), wars (Sarkees and Wayman 2010; Reiter,
  Stam, and Horowitz 2016), organized violence (Ralph Sundberg and Mihai
  Croicu 2016; Davies, Pettersson, and Öberg 2022), political violence
  (Raleigh et al. 2010), sanctions (Felbermayr et al. 2020), and
  international agreements (Kinne 2020; Owsiak, Cuttner, and Buck 2018),
  dispute resolution (Frederick, Hensel, and Macaulay 2017), and
  diplomacy (Moyer, Turner, and Meisel 2020; Sechser 2011).} Thanks to
improvements in natural language processing, more open-ended efforts
have begun to capture entire unstructured streams of events from
international news reports.\footnote{See Beieler et al. (2016); Boschee
  et al. (2015); Brandt et al. (2018); Grant et al. (2017); Li et al.
  (2021). On event extraction from images and social media see Zhang and
  Pan (2019) and Steinert-Threlkeld (2019).} How close these efforts are
to accurately measuring all or even most of what is essential in
international relations is an open empirical question, one for which we
provide new evidence here.

Our contribution is a high coverage ontology and event dataset for key
historical episodes in 20th and 21st-century international relations. We
develop a large, flexible ontology of international events with the help
of both human coders and natural language processing. We apply it
sentence-by-sentence to an unusually high-quality corpus of historical
narratives of international crises (Brecher 1999; Wilkenfeld and Brecher
2000; Brecher et al. 2021). The result is a new lower bound estimate of
how much actually happens between states during pivotal historical
episodes. We then develop several methods for objectively gauging how
well these event codings reconstruct the information contained in the
original narrative. We conclude by benchmarking our event codings
against several current state of the art event data collection efforts.
We find that existing systems produce sequences of events that do not
contain enough information to reconstruct the underlying historical
episode. The underlying fine-grained variation in international affairs
is unrecognizable through the lens of current quantification efforts.

@@

The second axis (columns) compares the very high ontological coverage of
ICBe to existing state of the art systems in production and with global
coverage. They begin with our contribution ICBe, alongside other
event-level datasets including CAMEO dictionary lookup-based systems
(Historical Phoenix (Althaus et al. 2019); ICEWS (Boschee et al. 2015);
Terrier (Grant et al. 2017)), the Militarized Interstate Disputes
Incidents dataset, and the UCDP-GED dataset (Davies, Pettersson, and
Öberg 2022; Sundberg and Melander 2013).\footnote{Other related datasets
  that insufficiently overlap ICBe's domain for comparison include BCOW
  (Leng and Singer 1988), WEIS (McClelland 1978), CREON (Hermann 1984),
  CASCON (Bloomfield and Moulton 1989), SHERFACS (Sherman 2000),
  Real-Time Phoenix (Brandt et al. 2018), and COfEE (Balali, Asadpour,
  and Jafari 2021) (see histories in Merritt (1994) and Schrodt and Hall
  (2006)).} The final set of columns compares episode-level datasets
beginning with the original ICB project (Brecher et al. 2021; Brecher
and Wilkenfeld 1982; Beardsley et al. 2020); the Militarized Interstate
Disputes dataset (Palmer et al. 2021; Gibler 2018), and the Correlates
of War (Sarkees and Wayman 2010). With the exception of large-scale
CAMEO dictionary-based systems, the existing state of the art
quantitative datasets ignore the vast majority of the information
content found in international relations.\footnote{See Balali, Asadpour,
  and Jafari (2021) for a recent review of ontological depth and
  availability of Gold Standard example text.}

@@

We find both the visualizations of macrostructure and head-to-head
comparisons of ICBe codings to the raw text to strongly support the
quality of ICBe, but as with recall, we seek a more objective detached
universal benchmark. Our proposed measure is a reconstruction task to
see whether our intended ontology can be recovered through only
unsupervised clustering of sentences they were applied to. Figure 4
shows the location of every sentence from the ICBe corpus in semantic
space as embedded using the same large language model as before, and the
median location of each ICBe event tag applied to those
sentences.\footnote{We preprocess sentences to replace named entities
  with a generic Entity token.} Labels reflect the individual leaves of
the ontology and colors reflect the higher level coarse branch nodes of
the ontology. If ICBe has high precision, substantively similar tags
ought to have been applied to substantively similar source text, which
is what we see both in two dimensions in the main plot and via
hierarchical clustering on all dimensions in the dendrogram along the
righthand side.\footnote{Hierarchical clustering on cosine similarity
  and with Ward's method.}

@@

We take the same approach here in Figure 5, selecting four specific
CAMEO event codings and checking how often they reflect a true
real-world event from the Crimea-Donbas synthetic narrative. We choose
four event types around key moments in the crisis. The start of the
crisis revolves around Ukraine backing out of a trade deal with the EU
in favor of Russia, but ``sign formal agreement'' events act more like a
topic detector with dozens of events generated by discussions of a
possible agreement but not the actual agreement which never
materialized. The switch is caught by the ``reject plan, agreement to
settle dispute'', but also continues for Viktor Yanukovych even after he
was removed from power because of articles retroactively discussing the
cause of his removal. Events for ``use conventional military force''
capture a threshold around the start of hostilities and who the
participants were but not any particular battles or campaigns. Likewise,
``impose embargo, boycott, or sanctions'' captures the start of waves of
sanctions and from who but are effectively constant as the news coverage
does not distinguish between subtle changes or additions. In sum,
dictionary-based methods on news corpora tend to have high recall
because they parse everything in the news, but for the same reason,
their specificity for most event types is too low to back out individual
chess-like sequencing that ICBe aims to record.

@@

We release several open-source products along with supporting code and
documentation to further advance the study of IR, event extraction, and
natural language processing. The first is the International Crisis
Behavior Events (ICBe) dataset, an event-level aggregation of what took
place during the crises identified by the ICB project. These data are
appropriate for statistical analysis of hard questions about the
sequencing of events (e.g.~escalation and de-escalation of
conflicts).\footnote{Using ICBe data Gannon (2022) finds that
  cross-domain crises are shorter and less violent than same-domain
  crises.} Second, we provide a coder-level disaggregation with multiple
codings of each sentence by experts and undergrads that allows for the
introduction of uncertainty and human interpretation of events. Further,
we release a direct mapping from the codings to the source text at the
sentence level as a new resource for natural language processing.
Finally, we provide a companion website that incorporates detailed
visualizations of all of the data introduced here at crisisevents.org.

@@

We thank the ICB Project and its directors and contributors for their
foundational work and their help with this effort. We make special
acknowledgment of Michael Brecher for helping found the ICB project in
1975, creating a resource that continues to spark new insights to this
day. We thank the many undergraduate coders. Thanks to the Center for
Peace and Security Studies and its membership for comments. Special
thanks to Rebecca Cordell, Philip Schrodt, Zachary Steinert-Threlkeld,
and Zhanna Terechshenko for their generous feedback. Thank you to the
cPASS research assistants: Helen Chung, Daman Heer, Syeda ShahBano Ijaz,
Anthony Limon, Erin Ling, Ari Michelson, Prithviraj Pahwa, Gianna Pedro,
Tobias Stodiek, Yiyi `Effie' Sun, Erin Werner, Lisa Yen, and Ruixuan
Zhang.

@@

This article's data, supplementary appendix, replication material, and
visualizations of every historical episode are available on the GitHub
repository
\href{https://urldefense.com/v3/__https://github.com/CenterForPeaceAndSecurityStudies/ICBEventData__;!!Mih3wA!WxDJtEczKfxGTh0S2Krunap8ReymFEL5iTWaSfOHeqlSdyfRx77zmjBSWO1OAm13$}{ICBEventData}
and through the companion website crisisevents.org.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-daviesOrganizedViolence19892022}{}}%
Davies, Shawn, Therése Pettersson, and Magnus Öberg. 2022. {``Organized
Violence 1989\textendash 2021 and Drone Warfare.''} \emph{Journal of
Peace Research} 59 (4): 593--610.

@@

@@
